---
title: Tidy Tuesday - Women's World Cup Results
author: Shelley Allen
date: '2019-07-12'
slug: tidy-tuesday
categories: []
tags: []
---

I thought a good place to start would be TidyTuesday data, this week it is Women's World Cup Results.

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  library(tidyverse)
```

```{r read-data, include=FALSE}
  wwc_outcomes <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-09/wwc_outcomes.csv")
  squads <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-09/squads.csv")
  codes <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-09/codes.csv")
  
  wwc_outcomes <- left_join(wwc_outcomes, codes, by = "team") %>%
    mutate(round = factor(round, levels = c("Final", "Semi Final", "Third Place Playoff", "Quarter Final", "Round of 16", "Group"), ordered = TRUE))
```

First, I had a look at the Country level stats to see who won the most games and got the most goals. I thought this would be interesting to see over the different Women's World Cups so I plotted a bar chart of the number of games won by each country facetted by year. Clearly Women's World Cup football is dominated by just a few teams but there are some surprising successes like Netherlands in 2019. I decided to plot a similar graph but for number of goals scored instead of games one which led me to create a function instead of repeating lots of code. 

```{r explore}
  total_wins <- wwc_outcomes %>%
    group_by(country) %>%
    summarise(total_wins = sum(win_status == "Won"), 
              total_goals = sum(score)) %>%
    arrange(desc(total_wins))
  
  country_wins <- wwc_outcomes %>% 
    group_by(country, year) %>%
    summarise(n_wins = sum(win_status == "Won"),
              n_goals = sum(score),
              best_round = min(round)) %>%
    left_join(total_wins) %>%
    arrange(year, desc(n_wins))
  
  country_year_plot <- function(df, column_name, column_label) {
    
    df %>%
      ggplot(aes(reorder(factor(country), -total_wins), !!column_name)) +
      geom_col(aes(fill = best_round)) +
      facet_grid(year ~ .) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
      labs(title = "The number of games won by each country in each Women's World Cup",
          x = "Country",
          y = column_label,
          fill = "Round achieved")
    }
  
  country_year_plot(country_wins, quo(n_wins), "Games won")
  country_year_plot(country_wins, quo(n_goals), "Goals scored")
```